---
title: "geoknife package"
author: "Jordan Read"
date: "`r format(Sys.time(), '%d %B, %Y')`"
output: 
  rmarkdown::html_vignette:
    toc: true
vignette: >
  %\VignetteEngine{knitr::rmarkdown}
  %\VignetteIndexEntry{geoknife package}
  \usepackage[utf8]{inputenc}
---


```{r setup, include=FALSE}
library(rmarkdown)
options(continue=" ")
options(width=60)
library(knitr)
library(geoknife)

```


#Introduction

The **geoknife** package ... 

#Installation
To install **geoknife**

Install package with dependencies:
```{r, eval=FALSE}
install.packages("geoknife", 
    repos = c("http://owi.usgs.gov/R"),
    dependencies = TRUE)
```

# geoknife web resources
## Geo Data Portal
## CSW (catalog service for the web)
## WFS (web feature service)
## WPS (web processing service)
# getting started
## geoknife concepts
## remote processing basics
## quick start guide
# gridded data (`fabric` argument for geoknife)
## `webgeom` class and details
## query for web available data
# spatial features (`stencil` argument for geoknife)
## `simplegeom` class and details
## `webgeom` class and details
## query for web available data
# web processing details (optional `knife` argument for geoknife)
## algorithm
## inputs
## url
## `wait`
## `email`
# geojob details
## `geojob` class and details
## `check` and related functions
## `cancel` geojob