---
title: "subset gridded data as geotiff"
author: "Jordan Read, Alison Appling"
date: "`r format(Sys.time(), '%d %B, %Y')`"
output: 
  rmarkdown::html_vignette:
    toc: true
    fig_width: 7
    fig_height: 6
vignette: >
  %\VignetteEngine{knitr::rmarkdown}
  %\VignetteIndexEntry{geoknife package}
  \usepackage[utf8]{inputenc}
---


```{r setup, include=FALSE}
library(rmarkdown)
options(continue=" ")
options(width=60)
library(knitr)
library(geoknife)

```


##Introduction  
`geoknife` has a number of output formats, but a new one that was added in version 1.1.0 of the package (after the initial release to CRAN) is the ability to output results as a zip file that contains a series of geotiffs for each timestep requested. This vignette is a brief introduction on how to do this using a few additional packages that are in the `Suggests` field in the description. 

### get the data
see the `geoknife` getting started vignette for more details, but for the purposes of this vignette, we need to get some data first and then plot it up.
```{r}
library(geoknife)
library(ggmap)
library(rasterVis)
library(ggplot2)

knife <- webprocess(algorithm = list('OPeNDAP Subset'="gov.usgs.cida.gdp.wps.algorithm.FeatureCoverageOPeNDAPIntersectionAlgorithm"))
fabric <- webdata(url='dods://opendap.larc.nasa.gov/opendap/SortByProduct/CERES/EBAF/Surface_Edition2.8/CERES_EBAF-Surface_Edition2.8_200003-201502.nc', 
                  variable="sfc_sw_down_all_mon", #Surface Shortwave Flux Down, Monthly Means, All-Sky conditions
                  times=c('2014-07-15','2014-07-15'))
stencil <- simplegeom(data.frame('point1'=c(-5,32), 'point2'=c(-90,-78))) # big 'ol chunk 'o data
job <- geoknife(stencil,fabric, knife, wait = TRUE, OUTPUT_TYPE="geotiff")
file <- download(job, destination = file.path(tempdir(), 'nasa_data.zip'), overwrite=TRUE)
unzip(file, exdir=file.path(tempdir(),'NASA'))
tiff.dir <- file.path(tempdir(),'NASA')
nasa <- raster(file.path(tiff.dir , dir(tiff.dir)))
world <- map_data("world2")
gplot(nasa, maxpixels = 5e5) + 
  geom_tile(aes(fill = value), alpha=1) +
  geom_polygon(data=world, aes(x=long, y=lat, group=group), color="navy", fill='transparent') +
  scale_fill_gradientn("Surface Shortwave Flux (W/m^2)", colours=rev(rainbow(5))) +
  coord_equal() + 
  theme_classic() + 
  theme(axis.line = element_blank(), axis.text.x = element_blank(), axis.text.y = element_blank(),
        axis.ticks = element_blank(), axis.title.x = element_blank(), axis.title.y = element_blank()) +
  scale_y_continuous(limits=c(nasa@extent@ymin, nasa@extent@ymax)) + 
  scale_x_continuous(limits=c(nasa@extent@xmin, nasa@extent@xmax))


